---
title: "Project1"
author: "Daniel Zinn"
date: "2024-12-23"
output: html_document
---

```{r setup, include=TRUE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```

## Reading and processing the data

Reading the given dataset activity.csv

```{r activity.csv , echo=TRUE}
activity_csv<-read.csv('activity.csv')

head(activity_csv)
```


### 1st plot "Total number of steps per day"

Processing the data by removing NAs, then grouping by the same date, and calculating the sum of the steps for each day.

```{r processing data , echo=TRUE}
steps_per_day<- activity_csv %>% filter(!is.na(steps)) %>%
  group_by(date) %>% 
  summarize(total_steps=sum(steps))

mean_steps<- mean(steps_per_day$total_steps)
median_steps<- median(steps_per_day$total_steps)

head(steps_per_day)
mean_steps
median_steps
```

The histogram plot war produced by the base library with the following code:

```{r histogram steps perday, echo=TRUE}
hist(steps_per_day$total_steps,main="Total steps taken per day",
     xlab = "Amount of steps",col = "lightblue")

abline(v=mean_steps,lwd=3, col="red")
abline(v=median_steps,lwd=3,lty=2, col="green")

legend("topright",
       legend=c(paste("Mean:",round(mean_steps,0)),
                paste("Median:",median_steps)),col=c("red","green"),
       lty=1,lwd=2,box.lwd =1,cex=0.8)
```

### Average daily activity pattern

Processing the dataset removing the NAs , grouping by Interval identifier and calculating the average across all days

```{r activity pattern plot, echo=TRUE}
five_minute<-activity_csv %>% 
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(average_steps=mean(steps))

head(five_minute)
```

Plotting using ggplot2 library:

```{r plotting average daily activity pattern}
ggplot(data=five_minute,aes(x=interval,y=average_steps))+
  geom_line(color="navy",lwd=1)+
  labs(title="Average steps per interval",
       y="average steps")+
  theme_minimal()
```

## Imputing missing values
for the next plot we are required to impute the missing values.

calculating the amount of rows with NA (because all NAs are in steps column there isnt an Intersection ): 

```{r amount of rows that contain NAs,echo=TRUE}
sum(is.na(activity_csv$interval))
sum(is.na(activity_csv$date))
sum(is.na(activity_csv$steps))
```

the strategy that we will use is calculating the mean of the steps for the same 5-minute interval and fill the missing data

```{r filling the missing data}
interval_means<- activity_csv %>% filter(!is.na(steps)) %>% 
  group_by(interval) %>% summarize(day_mean=mean(steps))

filled_activity_csv<-activity_csv
for (i in 1:length(filled_activity_csv[,1])) {
  if(is.na(filled_activity_csv[i,1])){
    interval_mean<-interval_means[interval_means$interval==filled_activity_csv[i,3],"day_mean"]
    filled_activity_csv[i,1]<-interval_mean
  }
}
```

We will let the reader compare the values of the first 5 variables

```{r compare,echo=TRUE}
head(interval_means)
head(filled_activity_csv)
```

creating histogram of total number of stepts taken per day
```{r histogram total number of steps per day,echo=TRUE}
steps_per_day2<- filled_activity_csv %>% filter(!is.na(steps)) %>%
  group_by(date) %>% 
  summarize(total_steps=sum(steps))

mean_steps<- mean(steps_per_day2$total_steps)
median_steps<- median(steps_per_day2$total_steps)

head(steps_per_day)
mean_steps
median_steps

hist(steps_per_day2$total_steps,main="Total steps taken per day",
     xlab = "Amount of steps",col = "lightblue")

abline(v=mean_steps,lwd=3, col="red")
abline(v=median_steps,lwd=3,lty=2, col="green")

legend("topright",
       legend=c(paste("Mean:",round(mean_steps,0)),
                paste("Median:",round(median_steps,0))),col=c("red","green"),
       lty=1,lwd=2,box.lwd =1,cex=0.8)

```

##activity patterns at: weekdays vs weekends
labeling the variables by weekend and weekdays
```{r proccessing the data,echo=TRUE}
week_days_ends <- filled_activity_csv %>%
  mutate(type = ifelse(weekdays(as.Date(date)) %in% c("Saturday", "Sunday"), "weekend", "weekday"))

week_days_ends<-week_days_ends %>% 
  filter(!is.na(steps)) %>%
  group_by(interval, type) %>%
  summarize(average_steps=mean(steps))

head(week_days_ends)
```

plotting the graph
```{r plotting weekends vs weekdays}
ggplot(data=week_days_ends,aes(x=interval,y=average_steps))+facet_wrap(~type,ncol=1)+
  geom_line(color="navy",lwd=1)+
  labs(title="Average steps per interval",
       y="average steps")+
  theme_minimal()
```