---
title: "Project1"
author: "Daniel Zinn"
date: "2024-12-23"
output: html_document
---

```{r setup, include=TRUE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```

## Reading and processing the data

Reading the given dataset activity.csv

```{r activity.csv , echo=TRUE}
activity_csv<-read.csv('activity.csv')

head(activity_csv)
```


### 1st plot "Total number of steps per day"

Processing the data by removing NAs, then grouping by the same date, and calculating the sum of the steps for each day.

```{r processing data , echo=TRUE}
steps_per_day<- activity_csv %>% filter(!is.na(steps)) %>%
  group_by(date) %>% 
  summarize(total_steps=sum(steps))

mean_steps<- mean(steps_per_day$total_steps)
median_steps<- median(steps_per_day$total_steps)

head(steps_per_day)
mean_steps
median_steps
```

The histogram plot war produced by the base library with the following code:

```{r histogram steps perday, echo=TRUE}
hist(steps_per_day$total_steps,main="Total steps taken per day",
     xlab = "Amount of steps",col = "lightblue")

abline(v=mean_steps,lwd=3, col="red")
abline(v=median_steps,lwd=3,lty=2, col="green")

legend("topright",
       legend=c(paste("Mean:",round(mean_steps,0)),
                paste("Median:",median_steps)),col=c("red","green"),
       lty=1,lwd=2,box.lwd =1,cex=0.8)
```

### Average daily activity pattern

Processing the dataset removing the NAs , grouping by Interval identifier and calculating the average across all days

```{r activity pattern plot, echo=TRUE}
five_minute<-activity_csv %>% 
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(average_steps=mean(steps))

head(five_minute)
```

Plotting using ggplot2 library:

```{r plotting average daily activity pattern}
ggplot(data=five_minute,aes(x=interval,y=average_steps))+
  geom_line(color="navy",lwd=1)+
  labs(title="Average steps per interval",
       y="average steps")+
  theme_minimal()
```

